%-----------------------------------------------------------------------------
%
%          PHYSICS  M.S.     THESIS
%          JUSTIN A. VASEL
%
%          This began as the template offered by the University of Minnesota, 
%          but I've made a few changes here and there...  
%
%          -->  acknowledge.tex
%
%-----------------------------------------------------------------------------



\chapterIntro{G}{raduate work is demanding.} It requires the student to utilize or develop many attributes; autonomy, dedication, discipline, and perseverance are a few. But, a successful post-graduate career also requires a strong support structure, on both a professional and personal level. 

There are many people that have earned my gratitude for their contribution to my success in graduate school. Dr. Alec Habig has been invaluable as a mentor. He taught me what it means to be an experimenter, and helped me stay afloat as I waded into the waters of particle physics for the first time.

Dr. Clarence Virtue and Dr. Michael Schumaker of Laurentian University in Ontario also deserve my thanks. Their kindness and guidance made joining the collaboration an educational and enjoyable experience. I wish to thank Dr. Mark Howe of the University of North Carolina at Chapel Hill who has answered many emails from me as I learned Objective-C and the ORCA source code.

I am thankful for my partner, Fernanda Psihas, who keeps me grounded when times are tough and challenges me to be the best that I can be. My closest friend, Matthew Reynolds, has always been a pillar of support before and during my graduate studies.

But, perhaps there are none more deserving of my gratitude than my parents, Steve Vasel and Jennifer Olson. My father was the first to teach me about the wonders of the Universe, instilling within me a deep and humble appreciation of everything it offers. My mother taught me the value of hard work and responsibility: the tools that have enabled me to pursue my passion. I would not be who I am or where I am today without their support; for that I will always be grateful.


%-----------------------------------------------------------------------------
%-----------------------------------------------------------------------------